Supplemental Information

### 2022-02-16 ms/manuscript.Rmd
Add SI pdf

---
title: "An open and continuously updated fern tree of life (FTOL)"
output: 
  bookdown::word_document2:
    pandoc_args: [ "--csl", "frontiers.csl"]
    toc: no
    number_sections: no
    reference_docx: Frontiers_Template.docx
editor_options: 
  chunk_output_type: console
params:
  doc_type: doc
---

@@

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')

# Load packages
library(conflicted)
library(rmarkdown)
library(bookdown)
library(targets)
library(assertr)
library(ape)
library(scales)
library(tidyverse)
library(rgnparser)
library(ggtree)
library(glue)
library(Polychrome)
library(patchwork)
library(magrittr)
library(deeptime)
library(ggrepel)
library(ggokabeito)
library(ggtext)
library(janitor)

# Resolve conflicts
conflict_prefer("filter", "dplyr")
conflict_prefer("extract", "magrittr")

# Load functions
source(here::here("R/functions.R"))
```

@@

tar_load(
	c(
		sanger_alignment,
		plastome_alignment,
		plastome_metadata_renamed,
		ppgi_taxonomy,
		plastid_calibration_dates,
		pteridocat,
		sanger_alignment_tbl,
		plastome_alignment_tbl,
		monophy_by_clade,
		plastome_tree,
		plastid_tree_dated,
		fossil_calibration_tips,
		treepl_cv_results),
	store = ftol_cache
)

@@

All sequences were downloaded from GenBank.
GenBank queries were formatted including terms "Polypodiopsida[ORGN]" to target ferns and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
Names including the terms "aff." or "cf.", hybrid formulas, and environmental DNA samples were excluded.
There is no formal definition of genomic vs. Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (=< `r number(7000)` bp) or plastome sequences (> `r number(7000)` bp) with the "[SLEN]" term.

@@

@@

The number of fern species sampled per locus in the Sanger dataset ranged from `r first(locus_total)$n` (*`r first(locus_total)$locus`*) to `r last(locus_total)$n`  (*`r last(locus_total)$locus`*).
`r sp_mult_loci$true$n` species (`r sp_mult_loci$true$percent`) were sampled for more than one locus.
Mean `r locus_summ$mean` ± `r locus_summ$sd` loci were sampled per species (all errors standard deviation unless otherwise mentioned).
The most frequent type of locus sampling per species was *rbcL* alone (`r locus_comb$rbcL$n` species).
The second most frequent type of locus sampling per species was all seven loci together (`r nth(locus_comb, length(locus_comb) - 1)$n` species). 
Locus sampling per species in the plastome dataset ranged from `r bb_locus_summ$min` to `r bb_locus_summ$max` (mean `r bb_locus_summ$mean` ± `r bb_locus_summ$sd` loci per species).
`r bb_with_all_loci$true$n` species (`r bb_with_all_loci$true$percent` of plastome species) included all 79 loci.

@@

```{r dna-stats}
# Calculate frequency of missing bases in each dataset overall
plastome_missing_freq <- base_freq_missing(plastome_alignment)
sanger_missing_freq <- base_freq_missing(sanger_alignment)

@@

The Sanger DNA alignment was `r number(ncol(sanger_alignment))` bp with `r percent(sanger_missing_freq, accuracy = 0.1)` missing data (missing bases or gaps) overall; rates of missing data by locus ranged from `r sanger_missing_by_gene %>% first() %>% extract2("missing")` (`r sanger_missing_by_gene %>% first() %>% extract2("gene")`) to `r sanger_missing_by_gene %>% last() %>% extract2("missing")` (`r sanger_missing_by_gene %>% last() %>% extract2("gene")`).
The plastome DNA alignment was `r number(ncol(plastome_alignment))` bp with `r percent(plastome_missing_freq, accuracy = 0.1)` missing data.

>>>>

---
title: "An open and continuously updated fern tree of life (FTOL)"
output: 
  bookdown::word_document2:
    pandoc_args: [ "--csl", "frontiers.csl"]
    toc: no
    number_sections: no
    reference_docx: Frontiers_Template.docx
editor_options: 
  chunk_output_type: console
params:
  doc_type: manual
---

@@

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')
# Load packages
source(here::here("R/ms_packages.R"))
```

```{r load-functions, eval = params$doc_type == "manual"}
# Load functions (only if running interactively)
source(here::here("R/functions.R"))
```

@@

tar_load(
	c(
		accs_exclude,
		fern_sanger_seqs_raw,
		raw_fasta_all,
		sanger_seqs_combined_filtered,
		sanger_seqs_rogues_removed,
		sanger_accessions_selection,
		plastome_seqs_combined_filtered,
		plastome_metadata_raw_all,
		plastome_metadata_renamed,
		sanger_alignment,
		plastome_alignment,
		plastome_metadata_renamed,
		ppgi_taxonomy,
		plastid_calibration_dates,
		pteridocat,
		sanger_alignment_tbl,
		plastome_alignment_tbl,
		monophy_by_clade,
		plastome_tree,
		plastid_tree_dated,
		fossil_calibration_tips,
		treepl_cv_results),
	store = ftol_cache
)

@@

All sequences were downloaded from GenBank.
GenBank queries were formatted including locus name, "Polypodiopsida[ORGN]", and "[PDAT]" to specify sequence date range (e.g., 1980/01/01[PDAT]:2021/12/12[PDAT]) so that a given query always results in the same set of sequences regardless of when the query is made.
Names including the terms "aff." or "cf.", hybrid formulas, and environmental DNA samples were excluded.
There is no formal definition of genomic vs. Sanger sequences in GenBank, so we used an empirical sequence length cutoff to distinguish between Sanger (=< `r number(7000)` bp) or plastome sequences (> `r number(7000)` bp) with the "[SLEN]" term.

@@

## GenBank mining

```{r genbank-stats}
# Helper function to count number of GenBank accessions in a dataset
count_accs <- function(data) {
	data %>%
	pull(accession) %>%
	n_distinct() %>%
	number()
}

# Manual list of accessions to exclude
n_accs_exclude <-
accs_exclude %>%
	pull(accession) %>%
	n_distinct() %>%
	number()

# Sanger ----

# Number of GenBank accessions resulting from inital query
n_accs_query <- count_accs(fern_sanger_seqs_raw)

# Number of GenBank accessions recovered with superCRUNCH
n_accs_extract <- count_accs(raw_fasta_all)

# Number of GenBank accessions after excluding manual rogues,
# names that couldn't be resolved
n_accs_first_filter <- count_accs(sanger_seqs_combined_filtered)

# Number of GenBank accessions after further excluding
# rogues based on BLAST
n_accs_second_filter <- count_accs(sanger_seqs_rogues_removed)

# Number of GenBank accessions in final selection
# first convert to long form
sanger_accessions_selection_long <-
 sanger_accessions_selection %>%
	select(contains("accession")) %>%
	pivot_longer(values_to = "accession", everything()) %>%
	filter(!is.na(accession))

# Count number of raw accessions in final selection
n_accs_raw_final <- count_accs(sanger_accessions_selection_long)

# Count number of accessions in final selection, split by locus
n_accs_final <- 
	sanger_accessions_selection_long %>%
	# Count accessions containing different loci as distinct
	mutate(
		name = str_remove_all(name, "accession_"),
		accession = paste(name, accession, sep = "_")) %>%
	count_accs()

# Analyze join by method
join_by_count <-
sanger_accessions_selection %>% 
	assert(not_na, join_by, species) %>%
	assert(is_uniq, species) %>%
	tabyl(join_by) %>%
	mutate(join_by = fct_reorder(join_by, n)) %>%
	mutate(
			percent = percent(percent, accuracy = 0.1),
			n = number(n, accuracy = 1)) %>%
	split(.$join_by)

# Plastome ----
# Don't include OG in calculations
n_plast_accs_query <-
	plastome_metadata_raw_all %>%
	left_join(
		select(plastome_metadata_renamed, accession, outgroup),
		by = "accession") %>%
	# excluded accessions lack outgroup info, but are all ferns
	mutate(outgroup = replace_na(FALSE)) %>%
	assert(not_na, outgroup) %>%
	filter(outgroup == FALSE) %>%
	count_accs()
n_plast_first_filter <-
	plastome_metadata_renamed %>%
  assert(not_na, outgroup) %>%
	filter(outgroup == FALSE) %>%
	count_accs()
n_accs_plast_final <- 
	plastome_seqs_combined_filtered %>%
	left_join(
		select(plastome_metadata_renamed, accession, outgroup),
		by = "accession") %>%
	assert(not_na, outgroup) %>%
	filter(outgroup == FALSE) %>%
	count_accs()
```

The initial GenBank query for the seven Sanger loci resulted in `r n_accs_query` accessions ("accessions" in this case includes some cases where a single accession contains multiple loci).
Extraction of target loci with superCRUNCH recovered `r n_accs_extract` accessions.
Manual inspection of the initial tree resulted in a list of `r n_accs_exclude` accessions to be excluded as rogues.
`r n_accs_first_filter` accessions were retained after excluding these rogues and species names that could not be resolved to an accepted name in the taxonomic database.
`r n_accs_second_filter` accessions were retained after excluding rogues identified by the all-by-all BLAST search.
The final selection of Sanger loci (one set of concatenated accessions per species or one accession per species if conditions for concatenation not met) included `r n_accs_raw_final` accessions (`r n_accs_final` sequences when counting distinct loci within each accession separately).
The most frequent method for joining accessions across loci was by `r last(join_by_count)$join_by` (`r last(join_by_count)$n` species; `r last(join_by_count)$percent`).
`r join_by_count$unjoined$n` species (`r join_by_count$unjoined$percent`) had accessions that could not be joined or only included one of the seven loci (`r s_table("join_by")`).

The initial GenBank query for fern plastomes resulted in `r n_plast_accs_query` accessions.
`r n_plast_first_filter` accessions were retained after excluding rogues and species names that could not be resolved.
Selection of the longest set of concatenated sequences per species resulted in `r n_accs_plast_final` accessions (species).

@@

The number of fern species sampled per locus in the Sanger dataset ranged from `r first(locus_total)$n` (*`r first(locus_total)$locus`*) to `r last(locus_total)$n`  (*`r last(locus_total)$locus`*).
`r sp_mult_loci$true$n` species (`r sp_mult_loci$true$percent`) were sampled for more than one locus.
Mean `r locus_summ$mean` ± `r locus_summ$sd` loci were sampled per species (all errors standard deviation unless otherwise mentioned).
The most frequent type of locus sampling per species was *rbcL* alone (`r locus_comb$rbcL$n` species) (`r s_figure("locus-upset")`).
The second most frequent type of locus sampling per species was all seven loci together (`r nth(locus_comb, length(locus_comb) - 1)$n` species). 
Locus sampling per species in the plastome dataset ranged from `r bb_locus_summ$min` to `r bb_locus_summ$max` (mean `r bb_locus_summ$mean` ± `r bb_locus_summ$sd` loci per species).
`r bb_with_all_loci$true$n` species (`r bb_with_all_loci$true$percent` of plastome species) included all 79 loci.

@@

```{r dna-stats}
# Calculate frequency of missing bases in each dataset overall
plastome_missing_freq <- base_freq_missing(plastome_alignment) %>%
	percent(accuracy = 0.1)
sanger_missing_freq <- base_freq_missing(sanger_alignment) %>%
	percent(accuracy = 0.1)

@@

The Sanger DNA alignment was `r number(ncol(sanger_alignment))` bp with `r sanger_missing_freq` missing data (missing bases or gaps) overall; rates of missing data by locus ranged from `r sanger_missing_by_gene %>% first() %>% extract2("missing")` (`r sanger_missing_by_gene %>% first() %>% extract2("gene")`) to `r sanger_missing_by_gene %>% last() %>% extract2("missing")` (`r sanger_missing_by_gene %>% last() %>% extract2("gene")`).
The plastome DNA alignment was `r number(ncol(plastome_alignment))` bp with `r plastome_missing_freq` missing data.

### 2022-02-16 ms/si.Rmd
Add SI pdf

---
title: "Supplemental Information"
output: 
  bookdown::word_document2:
    pandoc_args: [ "--csl", "frontiers.csl"]
    toc: no
    number_sections: no
    reference_docx: Frontiers_Template.docx
editor_options: 
  chunk_output_type: console
params:
  doc_type: doc
---

@@

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')

# Load packages
library(conflicted)
library(rmarkdown)
library(bookdown)
library(targets)
library(assertr)
library(ape)
library(scales)
library(tidyverse)
library(rgnparser)
library(ggtree)
library(glue)
library(Polychrome)
library(patchwork)
library(magrittr)
library(ggrepel)
library(ggokabeito)
library(ggforce)

# Resolve conflicts
conflict_prefer("filter", "dplyr")
conflict_prefer("extract", "magrittr")

# Load functions
source(here::here("R/ms_functions.R"))

# Specify percentage format
percent <- function(...) {scales::percent(...)}
number <- function(...) {scales::number(big.mark = ",", ...)}
```

```{r load-targets}
# Targets from FTOL workflow

@@

tar_load(
	c(
		sanger_sampling,
		monophy_by_clade,
		plastid_tree_dated),
	store = ftol_cache
)

@@

```{r age-biplot}
# Count number of tips in clades by taxonomy for double checking
n_tips_in_clades <-
	sanger_sampling %>%
	filter(outgroup == FALSE) %>%
	select(species, family, order) %>%
	pivot_longer(names_to = "level", values_to = "taxon", -species) %>%
	group_by(taxon) %>%
	summarize(number_tips_by_tax = n())

# Get ages for large clades
clade_ages <-
	sanger_sampling %>%
	select(family, order) %>%
	unique() %>%
	pivot_longer(names_to = "level", values_to = "taxon", everything()) %>%
	filter(!is.na(taxon)) %>%
	unique() %>%
	left_join(
		unique(select(monophy_by_clade, taxon, monophyly, mrca, number_tips)),
		by = "taxon") %>%
	mutate(across(c(mrca, number_tips), parse_number)) %>%
	assert(in_set(c("Yes", "Monotypic")), monophyly) %>%
	filter(monophyly == "Yes") %>%
	mutate(
		# Height is distance above root
		height = map_dbl(mrca, ~phytools::nodeheight(
			tree = plastid_tree_dated, node = .))
	) %>%
	mutate(
		# Age is the total length of the tree minus height
		age = max(phytools::nodeHeights(plastid_tree_dated)) - height
	) %>%
	# Double check that n_tips is as expected
	# (so the correct node is being used)
	left_join(n_tips_in_clades, by = "taxon") %>%
	verify(all(number_tips == number_tips_by_tax)) %>%
	select(-number_tips_by_tax)

clade_ages %>%
	filter(level == "family") %>%
	select(family = taxon, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
	select(matches("best|median")) %>%
	ggplot(aes(x = .panel_x, y = .panel_y)) +
	geom_point() +
	geom_autodensity(na.rm = TRUE) +
	facet_matrix(vars(everything()), layer.diag = 2, grid.y.diag = FALSE)
```

Figure 1: Biplot comparing crown ages of fern families (millions of years) across different studies.

>>>>

---
title: "Supplemental Information"
mainfont: Roboto-Regular.ttf
fontsize: 11pt
mainfontoptions: 
  - BoldFont=Roboto-Bold.ttf
  - ItalicFont=Roboto-Italic.ttf
  - BoldItalicFont=Roboto-BoldItalic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: default
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "frontiers.csl"]
header-includes: 
  - \usepackage{float} # keep figures from floating away
  - \usepackage{booktabs} # need for kable tables
  - \usepackage{colortbl} # need for kable tables
  - \makeatletter\renewcommand*{\fps@figure}{H}\makeatother # keep figures from floating away
  - \usepackage{fancyhdr} # enable header
  - \pagestyle{fancy}  # enable header
  - \fancypagestyle{plain}{\pagestyle{fancy}} # show header on first page
  - \renewcommand{\headrulewidth}{0pt} # delete line beneath header
  - \fancyhead[RE,RO]{Nitta \textit{et al}. Fern Tree of Life SI} # add header on right side
  - \fancyhead[LO,LE]{} # suppress automatic section header on left side
  - \cfoot{\thepage} # footer
editor_options: 
  chunk_output_type: console
params:
  doc_type: manual
---

@@

```{r setup-si, include = FALSE}
knitr::opts_chunk$set(
	echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')
# Load packages
library(ggupset)
library(kableExtra)
source(here::here("R/ms_packages.R"))
```

```{r load-functions-si, eval = params$doc_type == "manual"}
# Load functions (only if running interactively)
source(here::here("R/functions.R"))
```

```{r load-targets-si}
# Targets from FTOL workflow

@@

tar_load(
	c(
		sanger_alignment_tbl,
		sanger_sampling,
		sanger_accessions_selection),
	store = ftol_cache
)

@@

```{r load-captions, cache = FALSE}
# Source captions numbers from MS
# 
# So that caption numbers are in order of appearance in MS

# Run inline code when purling
options(knitr.purl.inline = TRUE)

# Make a temporary file to write out just inline R code from the SI Rmd
temp_file <- tempfile()

# Generate R script including inline R code from the ms Rmd
knitr::purl(here::here("ms/manuscript.Rmd"), output = temp_file)

# Trim this R script down to only functions that define figure and table captions
read_lines(temp_file) %>%
  magrittr::extract(
    str_detect(., "^figure\\(|^table\\(|^s_figure\\(|^s_table\\(|^figure_num\\(|^table_num\\(|^s_figure_num\\(|^s_table_num\\(")
  ) %>%
  unique %>%
  write_lines(temp_file)

# Source the code. This will load the caption functions in the right order.
source(temp_file, local = TRUE)

# Delete the temporary file
fs::file_delete(temp_file)
```

```{r join_by, results = "as-is"}
# Make table showing method of joining accessions across loci
sanger_accessions_selection %>% 
	left_join(select(sanger_sampling, species, outgroup), by = "species") %>%
	# FIXME why is this NA?
	# filter(is.na(outgroup)) %>%
	filter(outgroup == FALSE) %>%
	select(-outgroup) %>%
	assert(not_na, join_by, species) %>%
	assert(is_uniq, species) %>%
	tabyl(join_by) %>%
	arrange(n) %>%
	adorn_totals() %>%
	mutate(
		join_by = str_to_sentence(join_by) %>%
			str_replace_all("Monophy", "Monophyletic"),
		n = number(n),
		percent = case_when(
			join_by != "Total" ~ percent(percent) %>% 
				str_replace_all("%", "\\\\%"),
			TRUE ~ ""
		)
	) %>%
	rename(
		"Method" = join_by,
		`\\textit{n}` = n,
		`\\%` = percent
	) %>%
  kbl(booktabs = TRUE, escape = FALSE, format = "latex") %>%
  kable_styling(latex_options = c("striped", "HOLD_position"))
```

**`r s_table("join_by")`**. Methods used to join accessions across loci for Sanger dataset. Count is number of species.

\clearpage

```{r locus-upset, fig.width = 7, fig.height = 5}
# Make tibble with one row per locus per species
sanger_alignment_tbl %>%
	mutate(taxon = map(align_trimmed, rownames)) %>%
	select(locus = target, species = taxon) %>%
	unnest(cols = species) %>%
	# Exclude outgroups
	left_join(
		unique(select(sanger_sampling, species, outgroup)), by = "species") %>%
	filter(outgroup == FALSE) %>%
	select(-outgroup) %>%
	group_by(species) %>%
	summarize(locus = list(locus)) %>%
	ggplot(aes(x = locus)) +
	geom_bar() +
	labs(x = "Locus combination", y = "Count (num. species)") +
	scale_x_upset(n_intersections = 20)
```

**`r s_figure("locus-upset")`**.
Frequency of locus combinations by species, Sanger dataset.
Only top 20 combinations shown.
